---
title: "Exploring mtcars dataset"
author: "Valentin Goverdovsky"
date: "26 January 2016"
output: html_document
---

##Executive summary

This report explores **mtcars** dataset and in particular aims to answer the following two questions:

* Is an automatic or manual transmission better for MPG?
* Quantify the MPG difference between automatic and manual transmissions.

To analyse the dataset linear regression modelling methods are employed. Using these it was found that ...


##Exploring the dataset

First of all let's see if
```{r, echo=FALSE, fig.width=480, fig.height=480}
library(ggplot2)
data(mtcars)

g <- ggplot(mtcars, aes(x = factor(am), y = mpg)) + 
     geom_boxplot(aes(fill=factor(am))) + 
     labs(x = 'Transmission', y = 'MPG') + 
     scale_fill_discrete(name='Transmission Type',
                         breaks=c('0', '1'),
                         labels=c('Automatic', 'Manual')) +
     theme(axis.ticks = element_blank(), axis.text.x = element_blank())
g
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
